---
title: "First analyses"
author: "Kalle Stoffers"
date: "2026-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

rm(list=ls())

library(tidyverse)
library(sf)
library(tmap)
library(spdep)
library(lme4)
library(lmerTest)
library(performance)
library(see)
library(fixest)
library(clubSandwich)


df <- st_read("./data/df_analysis.gpkg")
```

## Plots

Our dependent variable is the percentage of electricity connections in a postal code area that have a yearly net negative electricity usage. We see that most areas have a low proportion of net producers and that over time, this percentage has increased. (Data from 2021 and 2020 is missing because there was no CBS data from 2019 and 2020, but this should become available soon.)

```{r}

hist(df$teruglevering)

df |>
  group_by(year) |>
  summarise(yearly_avg = mean(teruglevering, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = yearly_avg)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(min(df$year), max(df$year), by = 2))+
  theme_minimal()
```

Let's plot the percentage of net producers on a map. First, using ggplot:

```{r}
#on the street level
df |>
  filter(year == 2024) |>
  ggplot(
    aes(fill = teruglevering)
  ) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "plasma", name = "% net producers") +
  theme_minimal()

#neighborhood average:
df |>
  filter(year == 2024) |>
  ggplot(
    aes(fill = nb_pv)
  ) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "plasma", name = "% net producers") +
  theme_minimal()

```

Now we do the same using tmap. We see that some postal code areas are much larger than others. Furthermore, some areas are only available in some years.

```{r}
tmap_mode("view")

tm_shape(df |> filter (year == 2024)) +
  tm_fill("teruglevering")

tm_shape(df |> filter (year == 2023)) +
  tm_fill("teruglevering")

tm_shape(df |> filter (year == 2022)) +
  tm_fill("teruglevering")

tm_shape(df |> filter (year == 2019)) +
  tm_fill("teruglevering")

tm_shape(df |> filter (year == 2018)) +
  tm_fill("teruglevering")

tm_shape(df |> filter (year == 2017)) +
  tm_fill("teruglevering")

tm_shape(df |> filter (year == 2016)) +
  tm_fill("teruglevering")
```

Calculate spatial autocorrelation over time, using "queen" contiguity to define the neighbors. There seems to be a decent amount of spatial autocorrelation, and it increases somewhat over time.

```{r}
years <- c(2016:2019, 2022:2024)
moran_results_list <- list()

for (yr in years) {
  
  # Filter for the specific year
  df_yr <- df |> filter(year == yr)
    
    # Create neighbor list
    nb <- poly2nb(df_yr)
    
    # Create weights list
    lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
    
    # Run Moran's I test and store it in the list using the year as the name
    moran_results_list[[as.character(yr)]] <- moran.test(df_yr$teruglevering, 
                                                         lw, 
                                                         zero.policy = TRUE)
    
}

str(moran_results_list)
```

Now let's calculate the spatial autocorrelation based on a weight matrix defined by proximity instead of contiguity. Very low Moran's I.

```{r, eval=FALSE}
years <- c(2016:2019, 2022:2024)
moran_results_list <- list()

for (yr in years) {
  
  # Filter for the specific year
  df_yr <- df |> filter(year == yr)
  
  # Extract centroids and coordinates
  centroids <- st_centroid(df_yr)
  coords <- st_coordinates(centroids)
  
  # Compute full distance matrix between all centroids
  D <- as.matrix(dist(coords))
  
  # Compute proximity weights: W = 1 - (D / max(D))^2
  max_D <- max(D)
  W <- 1 - (D / max_D)^2
  
  # Set diagonal to 0 (no self-neighbors)
  diag(W) <- 0
  
  # Convert to listw object
  lw <- mat2listw(W, style = "W", zero.policy = TRUE)
  
  # Run Moran's I test
  moran_results_list[[as.character(yr)]] <- moran.test(df_yr$teruglevering,
                                                       lw,
                                                       zero.policy = TRUE)
}

str(moran_results_list)

```

As a preliminary analysis, let's look at the relationship between the percentage of net producers in the neighborhood at t-1 and the percentage of net producers at t0. It looks like there is an S-curve (sort of), in line with the literature on innovation adoption.

```{r}
df |>
  ggplot(aes(x = nb_pv, y = teruglevering)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  theme_minimal() +
  xlab("Neighborhood net producers at t-1") +
  ylab("Net producers at t=0")
```

## First models

In RSiena I couldn't find any effect that tested the strength of influence of similar others compared to dissimilar others, so I for now I have opted for panel regression analysis where I interact the effect neighborhood net producers (nb_pv) with neighborhood demographic diversity, income, etc.

In the models with see that neighborhood pv-adoption has a positive effect on pv-adoption, following a logistic curve.

```{r}
df <- df|>
  mutate(time = year - 2016)

#null-model, ICC = 34.7%. 
#Most of the variation is within-streets over time, interesting!
model0 <- lmer(teruglevering ~ (1 | postcode_van), REML = F, data = df)
summary(model0)
149.9 / (149.9 + 282.3)

#model1 with random interept and fixed effect of time
model1 <- lmer(teruglevering ~ (1 | postcode_van) + time, REML = F, data = df)
summary(model1)

#model2 with random effect of time
model2 <- lmer(teruglevering ~ (1 + time | postcode_van) + time, REML = F, data = df)
summary(model2)

#model3 with random  intercept and random effect of time and fixed effect of time and nb_pv
model3 <- lmer(teruglevering ~ (1 + time | postcode_van) + time + nb_pv, REML = F, data = df)
summary(model3)

#model4 with random intercept and random effect of time and nb_pv
model4 <- lmer(teruglevering ~ (1 + time + nb_pv | postcode_van) + time + nb_pv, REML = F, data = df)
summary(model4)

#model5
df <- df |> 
  mutate(nb_pv_sq = nb_pv^2,
         nb_pv_log = 1 / (1 + exp(-nb_pv))
         )

#with nb_pv squared
model5_1 <- lmer(teruglevering ~ (1 + time + nb_pv | postcode_van) + time + nb_pv + nb_pv_sq, REML = F, data = df)
summary(model5_1)

#with logistic function of nb_pv ("S-curve")
#higher significance and lower BIC so we stick with nb_pv_log
model5_2 <- lmer(teruglevering ~ (1 + time + nb_pv | postcode_van) + time + nb_pv + nb_pv_log, REML = F, data = df)
summary(model5_2)

```

Because of the different scales, we standardize all variables. We add the CBS demographic information as interaction variables on the nb_pv variable.

```{r}
model6 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(solar_cost)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(stedelijkheid)
                + scale(percentage_koopwoningen))
               + time 
               + scale(nb_pv)
               + scale(nb_pv_log), REML = F, data = df)
summary(model6)

#check for multicollinearity
multicollinearity(model6)

#moderate correlations (VIF >5) for interaction terms with #percentage_laag_inkomen_huishouden, solar_cost, omgevingsadressendichtheid, #stedelijkheid. So we chose only one density measure (omgevingsadressendichtheid),
#and remove solar_cost, this probably correlates too much with time. 

model7 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen)
                + time)
               + scale(nb_pv)
               + scale(nb_pv_log), REML = F, data = df)
summary(model7)
multicollinearity(model7)


#time:nb_pv also too much multicollinearity so we leave it out
model8 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
               + time
               + scale(nb_pv)
               + scale(nb_pv_log), REML = F, data = df)
summary(model8)

png("check_model8_output.png", width = 2000, height = 1600, res = 150)
plot(performance::check_model(model8))
dev.off()


#adding non-linear interaction terms
df <- df |>
  mutate(gemiddelde_woz_waarde_woning_sq = gemiddelde_woz_waarde_woning^2,
         electricity_price_sq = electricity_price^2)

model9 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(gemiddelde_woz_waarde_woning_sq)
                + scale(electricity_price)
                + scale(electricity_price_sq)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
               + time
               + scale(nb_pv)
               + scale(nb_pv_log), REML = F, data = df)
summary(model9)

png("check_model9_output.png", width = 2000, height = 1600, res = 150)
plot(performance::check_model(model9))
dev.off()

#too much multicollinearity for non-linear interaction terms, have to leave them #out
multicollinearity(model9)
```

Add control variables, age and ethnicity

```{r}
model10 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
               + time 
               + scale(nb_pv)
               + scale(nb_pv_log)
               + scale(age_avg)
               + scale(percentage_geb_nederland_herkomst_nederland), 
               REML = F, data = df)
summary(model10)

png("check_model10_output.png", width = 2000, height = 1600, res = 150)
plot(performance::check_model(model10))
dev.off()


#let's try a model with only areas with home ownership >= 50%
summary(df$percentage_koopwoningen)

df_own <- df |> 
  filter(percentage_koopwoningen >= 50)

#won't converge, too small a sample size?
model11 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
               + time 
               + scale(nb_pv)
               + scale(nb_pv_log)
               + scale(age_avg)
               + scale(percentage_geb_nederland_herkomst_nederland), 
               REML = F, data = df_own)

```

Now using a model with REML = T. Interpretation of results, top to bottom:

-   *Neighborhood PV-adoption has a positive effect on PV adoption (logistic function)*

-   More high income households, more PV

-   Higher average property value, more PV

-   Higher electricity price, more PV

-   More home ownership, more PV

-   PV-adoption has increased over time

-   Higher average age, less PV

-   *The effect of neighborhood PV is weaker with more high income households*

-   *No difference in the effect of neighborhood PV based on percentage of low income households*

-   *The effect of neighborhood PV is weaker with a higher average property value*

-   *No difference in the effect of neighborhood PV based on electricity price*

-   *No difference in the effect of neighborhood PV based on economic diversity*

-   *The effect of neighborhood PV is weaker with a higher ethnic diversity*

-   *The effect of neighborhood PV is weaker with a higher age diversity*

-   *No difference in the effect of neighborhood PV based on density*

-   *The effect of neighborhood PV is stronger with more home ownership*

```{r}
model12 <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
               + time 
               + scale(nb_pv)
               + scale(nb_pv_log)
               + scale(age_avg)
               + scale(percentage_geb_nederland_herkomst_nederland), 
               REML = T, data = df)
summary(model12)

png("check_model12_output.png", width = 2000, height = 1600, res = 150)
plot(performance::check_model(model12))
dev.off()

#check more assumptions
#autocorrelation and heteroscedasticity present
check_collinearity(model12)
check_autocorrelation(model12)
check_heteroscedasticity(model12)

#using cluster-robust SEs
model12_rob_res <- coef_test(model12, vcov = "CR2", 
          cluster = model.frame(model12)$postcode_van)

view(model12_rob_res)


```

Baseline model without any interaction terms to see how much of the slope variance in nb_pv we explained

```{r}
model_bl <- lmer(scale(teruglevering) ~ (1 + time + scale(nb_pv) | postcode_van) 
               + scale(nb_pv)
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen)
               + time 
               + scale(nb_pv)
               + scale(nb_pv_log)
               + scale(age_avg)
               + scale(percentage_geb_nederland_herkomst_nederland), 
               REML = T, data = df)
summary(model_bl)

# variance in the random slope of scale(nb_pv) is actually lower than in the #interaction model???


```

Trying fixed effect model.

```{r}
library(fixest)

model_fe <- feols(
  scale(teruglevering) ~ 
    scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
     + time 
     + scale(nb_pv)
     + scale(nb_pv_log)
     + scale(age_avg)
     + scale(percentage_geb_nederland_herkomst_nederland) 
  | postcode_van, 
  data = df)
  
summary(model_fe)

#autocorrelation and heteroscedasticity so we should use cluster robust SEs
check_collinearity(model_fe)
check_autocorrelation(model_fe)
check_heteroscedasticity(model_fe)


model_fe_rob <- feols(
  scale(teruglevering) ~ 
    scale(nb_pv) * (
                + scale(percentage_hoog_inkomen_huishouden)
                + scale(percentage_laag_inkomen_huishouden)
                + scale(gemiddelde_woz_waarde_woning)
                + scale(electricity_price)
                + scale(economic_diversity)
                + scale(ethnic_diversity)
                + scale(age_diversity)
                + scale(omgevingsadressendichtheid)
                + scale(percentage_koopwoningen))
     + time 
     + scale(nb_pv)
     + scale(nb_pv_log)
     + scale(age_avg)
     + scale(percentage_geb_nederland_herkomst_nederland) 
  | postcode_van,
  cluster = "postcode_van",
  data = df)
  
summary(model_fe_rob)

```

## Questions and Future

-   Non-linear interactions?
-   How much variance in the slope of nb_pv do I explain?
-   Decide on model, MGM vs FE?
    -   I think MGM might be preferable, since some variables might not vary enough year by year.
-   Impute missing data with spatial interpolation?
-   Neighborhood definition based on distance instead of 500m x 500m squares?
    -   
