---
title: "Journal 2"
author: "Kalle Stoffers"
date: "2026-03-04"
output: html_document
---

Last compiled on `r format(Sys.time(), '%B, %Y')`

<br>

------------------------------------------------------------------------

## First look at CBS data

After having assembled the solar panel data, it is time to connect it to the demographic characteristics of the area.

First, we load the relevant packages

```{r eval = FALSE, message=FALSE}
rm(list=ls())

library(tidyverse)
library(sf)
```

Age, ethnic background, percentage of home ownership, average property value and electricity usage is known at the postal code level, so we import the CBS data per PC6 postal code. These data can be downloaded here: <https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/gegevens-per-postcode>

```{r eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
#load data
cbs_pc6_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_pc6_23 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_pc6_22 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_pc6_21 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2021_vol.gpkg") |>
  mutate (year = 2021)

#data from 2020 and 2019 is unavailable at this moment but will be available before the 1st of May

cbs_pc6_18 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2018-v3/CBS_PC6_2018_v3.shp") |>    mutate (year = 2018)
cbs_pc6_17 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2017-v3/CBS_PC6_2017_v3.shp") |>
  mutate (year = 2017)
cbs_pc6_16 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2016-v2/CBS_PC6_2016_v2.shp") |>
  mutate (year = 2016)
cbs_pc6_15 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2015-v2/CBS_PC6_2015_v2.shp") |>
  mutate (year = 2015)


#merge and NAs
cbs_pc6 <- bind_rows(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22) |>
  arrange (postcode6, year) |>
  relocate (postcode6, year, geom)|>
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995)))
```

Some streets in the PV data have multiple postal codes:

```{r eval=FALSE}
#load pv data and add id
pv <- readRDS("./data/pv_data.Rds") |>
  filter(year >= 2022) |>
  mutate(id = row_number()) |>
  relocate(id)

#check for multiple postal codes per street
meerdere_postcodes <- pv |> 
  mutate(same_pc = if_else(POSTCODE_VAN == POSTCODE_TOT, 1, 0)) |>
  filter(same_pc == 0) |>
  distinct(STRAATNAAM, POSTCODE_VAN, POSTCODE_TOT, .keep_all = TRUE)

```

We select the CBS postal codes which match the streets from the PV data.

```{r eval=FALSE, results = 'hide'}
cbs_pc6_sel <- cbs_pc6 |>
  inner_join(
    pv, 
    by = join_by(
      year == year,
      postcode6 >= POSTCODE_VAN,
      postcode6 <= POSTCODE_TOT
    )
  ) |>
  relocate(postcode6, geom, year, id, POSTCODE_VAN, POSTCODE_TOT, STRAATNAAM, X.Leveringsrichting)

```

Now we collapse the multiple rows into a single row per street in the PV data.

```{r eval=FALSE}
df <- cbs_pc6_sel |>
  group_by(id) |>
  summarise(
    
    geometry = sf::st_union(geom),
    
    across(
      c(
        POSTCODE_VAN, 
        POSTCODE_TOT, 
        STRAATNAAM, 
        year,
        X.Leveringsrichting
        ), 
           first),
    
    across(
      c(
        percentage_geb_nederland_herkomst_nederland,
        percentage_geb_nederland_herkomst_overig_europa,
        percentage_geb_nederland_herkomst_buiten_europa,
        percentage_geb_buiten_nederland_herkomst_europa,
        percentage_geb_buiten_nederland_herkmst_buiten_europa
      ),
      ~ weighted.mean(.x, aantal_inwoners, na.rm = TRUE)
    ),
    
    gemiddelde_huishoudensgrootte = 
      weighted.mean(gemiddelde_huishoudensgrootte,
                    aantal_part_huishoudens, 
                    na.rm = TRUE),
    
    across(
      c(
        percentage_koopwoningen, 
        percentage_huurwoningen, 
        gemiddelde_woz_waarde_woning, 
        gemiddeld_elektriciteitsverbruik_woning
        ), 
           ~weighted.mean(.x, aantal_woningen, na.rm = TRUE)
      ),
    
    across(
      c(
        aantal_inwoners,
        aantal_mannen,
        aantal_vrouwen,
        aantal_inwoners_0_tot_15_jaar,
        aantal_inwoners_15_tot_25_jaar,
        aantal_inwoners_25_tot_45_jaar,
        aantal_inwoners_45_tot_65_jaar,
        aantal_inwoners_65_jaar_en_ouder,
        aantal_part_huishoudens,
        aantal_woningen,
        aantal_huurwoningen_in_bezit_woningcorporaties,
        aantal_niet_bewoonde_woningen,
        aantal_personen_met_uitkering_onder_aowlft
      ),
      ~ sum(.x, na.rm = TRUE)
      )
  ) |>
  rename(postcode_van = POSTCODE_VAN,
         postcode_tot = POSTCODE_TOT,
         straatnaam = STRAATNAAM,
         leveringsrichting = X.Leveringsrichting)
```

Income and urbanity/density are only available per 500m x 500m square. This data can be downloaded here: [https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/kaart-van-500-meter-bij-500-meter-met-statistieken](#0){.uri}

```{r eval=FALSE, message = FALSE}
cbs_500_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_500_23 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2023_v2/cbs_vk500_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_500_22 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2022_vol/cbs_vk500_2022_vol.gpkg") |>
  mutate (year = 2022)

cbs_500 <- bind_rows(cbs_500_24, cbs_500_23, cbs_500_22) |>
  
  select(crs28992res500m, gemiddeld_inkomen_huishouden, percentage_laag_inkomen_huishouden, percentage_hoog_inkomen_huishouden, omgevingsadressendichtheid, stedelijkheid, geom, year)|>
  
  relocate(crs28992res500m, geom, year, gemiddeld_inkomen_huishouden, percentage_laag_inkomen_huishouden, percentage_hoog_inkomen_huishouden, omgevingsadressendichtheid, stedelijkheid) |>
  
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995)))
```

Now we join the squares with the postal code areas based on the largest overlapping area.

```{r eval =FALSE}
#match the squares belonging to the postal code area and keep only the square id
df <- df |>
  st_join(cbs_500,join = st_intersects, left = T, largest = T) |>
  select(-c(year.y, gemiddeld_inkomen_huishouden, percentage_laag_inkomen_huishouden, percentage_hoog_inkomen_huishouden, omgevingsadressendichtheid, stedelijkheid))

#match the square attributes in the right year
df <- df |>
  left_join (st_drop_geometry(cbs_500), by = c("crs28992res500m", "year.x" = "year")) |>
  rename(year = year.x)

```

```{r eval=FALSE}
df <- df |> 
  mutate(across(where(is.numeric), ~na_if(., NaN)))
```

Saving the data!

```{r eval=FALSE}
saveRDS(df, file = "./data/merged_data.Rds")

```
