---
title: "Journal 2"
author: "Kalle Stoffers"
date: "2026-03-04"
output: html_document
---

Last compiled on `r format(Sys.time(), '%B, %Y')`

<br>

------------------------------------------------------------------------

## First look at CBS data

After having assembled the solar panel data, it is time to connect it to the demographic characteristics of the area.

First, we load the relevant packages

```{r eval = FALSE, message=FALSE}
rm(list=ls())

library(tidyverse)
library(sf)
```

Most of the data we want is known at the postal code level, so we import the CBS data per PC6 postal code. These data can be downloaded here: <https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/gegevens-per-postcode>. The data from 2020 and 2019 is unavailable at this moment but should be available before the 1st of May. The data from 2021-2024 and 2015-2018 are named a bit differently so we first rename and select the relevant columns before joining.

```{r eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
#load data
cbs_pc6_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_pc6_23 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_pc6_22 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_pc6_21 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2021_vol.gpkg") |>
  mutate (year = 2021) |>
  rename (percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)

cbs_pc6_2124 <- bind_rows(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22, cbs_pc6_21) |>
  select (- c(10, 17:20, 23:31, 37, 40:131))

rm(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22, cbs_pc6_21)
gc()

cbs_pc6_18 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2018-v3/CBS_PC6_2018_v3.shp") |>  
  mutate (year = 2018) |>
  rename (Postcode = PC6)
cbs_pc6_17 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2017-v3/CBS_PC6_2017_v3.shp") |>
  mutate (year = 2017)
cbs_pc6_16 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2016-v2/CBS_PC6_2016_v2.shp") |>
  mutate (year = 2016)
cbs_pc6_15 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2015-v2/CBS_PC6_2015_v2.shp") |>
  mutate (year = 2015)

cbs_pc6_1518 <- bind_rows(cbs_pc6_18, cbs_pc6_17, cbs_pc6_16, cbs_pc6_15) |>
  rename (
    postcode6 = Postcode,
    aantal_inwoners = INWONER,
    aantal_mannen = MAN,
    aantal_vrouwen = VROUW,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    aantal_part_huishoudens = AANTAL_HH,
    gemiddelde_huishoudensgrootte = GEM_HH_GR,
    aantal_woningen = WONING,
    percentage_koopwoningen = P_KOOPWON, 
    percentage_huurwoningen = P_HUURWON,
    aantal_huurwoningen_in_bezit_woningcorporaties = WON_HCORP,
    aantal_niet_bewoonde_woningen = WON_NBEW,
    gemiddelde_woz_waarde_woning = WOZWONING, 
    gemiddeld_elektriciteitsverbruik_woning = G_ELEK_WON,
    aantal_personen_met_uitkering_onder_aowlft = UITKMINAOW,
    geom = geometry
  ) |>
  select (- c(10, 15:18, 21:29, 35, 39:130))
  
rm(cbs_pc6_18, cbs_pc6_17, cbs_pc6_16, cbs_pc6_15)
gc()

#join and set NAs
cbs_pc6 <- bind_rows(cbs_pc6_2124, cbs_pc6_1518) |>
  relocate (postcode6, year, geom)|>
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995))) 

rm(cbs_pc6_2124, cbs_pc6_1518)
gc()

```

Some streets in the PV data have multiple postal codes:

```{r eval=FALSE}
#load pv data and add id
pv <- readRDS("./data/pv_data.Rds") |>
  filter(year >= 2015) |>
  mutate(id = row_number()) |>
  relocate(id)

#check for multiple postal codes per street
meerdere_postcodes <- pv |> 
  mutate(same_pc = if_else(POSTCODE_VAN == POSTCODE_TOT, 1, 0)) |>
  filter(same_pc == 0) |>
  distinct(STRAATNAAM, POSTCODE_VAN, POSTCODE_TOT, .keep_all = TRUE)

```

We select the CBS postal codes which match the streets from the PV data.

```{r eval=FALSE, results = 'hide'}
cbs_pc6_westland <- cbs_pc6 |>
  inner_join(
    pv, 
    by = join_by(
      year == year,
      postcode6 >= POSTCODE_VAN,
      postcode6 <= POSTCODE_TOT
    )
  ) |>
  relocate(postcode6, geom, year, id, POSTCODE_VAN, POSTCODE_TOT, STRAATNAAM, X.Leveringsrichting)

rm(cbs_pc6, meerdere_postcode)
gc()

```

Now we collapse the multiple rows into a single row per street in the PV data.

```{r eval=FALSE}
df <- cbs_pc6_westland |>
  group_by(id) |>
  summarise(
    
    geometry = sf::st_union(geom),
    
    across(
      c(
        POSTCODE_VAN, 
        POSTCODE_TOT, 
        STRAATNAAM, 
        year,
        X.Leveringsrichting
        ), 
           first),
    
    across(
      c(
        percentage_geb_nederland_herkomst_nederland,
        percentage_geb_nederland_herkomst_overig_europa,
        percentage_geb_nederland_herkomst_buiten_europa,
        percentage_geb_buiten_nederland_herkomst_europa,
        percentage_geb_buiten_nederland_herkmst_buiten_europa,
        percentage_westerse_migr_achtergr,
        percentage_niet_westerse_migr_achtergr
        ),
      ~ weighted.mean(.x, aantal_inwoners, na.rm = TRUE)
    ),
    
    gemiddelde_huishoudensgrootte = 
      weighted.mean(gemiddelde_huishoudensgrootte,
                    aantal_part_huishoudens, 
                    na.rm = TRUE),
    
    across(
      c(
        percentage_koopwoningen, 
        percentage_huurwoningen, 
        gemiddelde_woz_waarde_woning, 
        gemiddeld_elektriciteitsverbruik_woning
        ), 
           ~weighted.mean(.x, aantal_woningen, na.rm = TRUE)
      ),
    
    across(
      c(
        aantal_inwoners,
        aantal_mannen,
        aantal_vrouwen,
        aantal_inwoners_0_tot_15_jaar,
        aantal_inwoners_15_tot_25_jaar,
        aantal_inwoners_25_tot_45_jaar,
        aantal_inwoners_45_tot_65_jaar,
        aantal_inwoners_65_jaar_en_ouder,
        aantal_part_huishoudens,
        aantal_woningen,
        aantal_huurwoningen_in_bezit_woningcorporaties,
        aantal_niet_bewoonde_woningen,
        aantal_personen_met_uitkering_onder_aowlft
      ),
      ~ sum(.x, na.rm = TRUE)
      )
  ) |>
  rename(postcode_van = POSTCODE_VAN,
         postcode_tot = POSTCODE_TOT,
         straatnaam = STRAATNAAM,
         leveringsrichting = X.Leveringsrichting)
```

Income and urbanity/density are only available per 500m x 500m square. This data can be downloaded here: [https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/kaart-van-500-meter-bij-500-meter-met-statistieken](#0){.uri}

```{r eval=FALSE, message = FALSE}
cbs_500_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_500_23 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2023_v2/cbs_vk500_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_500_22 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2022_vol/cbs_vk500_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_500_21 <- st_read(dsn = "C:/Users/kalle/Downloads/2024-cbs_vk500_2021_vol/cbs_vk500_2021_vol.gpkg") |>
  mutate (year = 2021) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_20 <- st_read(dsn = "C:/Users/kalle/Downloads/2023-cbs_vk500_2020_vol/cbs_vk500_2020_vol.gpkg") |>
  mutate (year = 2020) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_19 <- st_read(dsn = "C:/Users/kalle/Downloads/2023-cbs_vk500_2019_vol2/cbs_vk500_2019_vol2.gpkg") |>
  mutate (year = 2019) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_18 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2018.gpkg") |>
  mutate (year = 2018) |>
  rename(crs28992res500m = c28992r500, 
         percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_17 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2017.gpkg") |>
  mutate (year = 2017) |>
  rename(crs28992res500m = c28992r500,
         percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)

cbs_500_16 <- st_read(dsn = "C:/Users/kalle/Downloads/2019-cbs_vk500_2016_v2/CBS_VK500_2016_v2.shp") |>
  mutate (year = 2016) |>
  rename(
    crs28992res500m = C28992R500,
    aantal_inwoners = INWONER,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    percentage_laag_inkomen_huishouden = P_LINK_HH, 
    percentage_hoog_inkomen_huishouden = P_HINK_HH, 
    omgevingsadressendichtheid = OAD, 
    stedelijkheid = STED, 
    geom = geometry
  )

cbs_500_15 <- st_read(dsn = "C:/Users/kalle/Downloads/2019-cbs_vk500_2015_v2/CBS_VK500_2015_v2.shp") |>
  mutate (year = 2015) |>
  rename(
    crs28992res500m = C28992R500,
    aantal_inwoners = INWONER,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    percentage_laag_inkomen_huishouden = P_LINK_HH, 
    percentage_hoog_inkomen_huishouden = P_HINK_HH, 
    omgevingsadressendichtheid = OAD, 
    stedelijkheid = STED, 
    geom = geometry
  )

cbs_500 <- bind_rows(cbs_500_24, cbs_500_23, cbs_500_22, cbs_500_21, cbs_500_20, cbs_500_19, cbs_500_18, cbs_500_17, cbs_500_16, cbs_500_15) 

rm(cbs_500_24, cbs_500_23, cbs_500_22, cbs_500_21, cbs_500_20, cbs_500_19, cbs_500_18, cbs_500_17, cbs_500_16, cbs_500_15)
gc()

cbs_500 <- cbs_500 |>
  
  select(crs28992res500m, 
          aantal_inwoners,
          aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland,
          percentage_laag_inkomen_huishouden,
          percentage_hoog_inkomen_huishouden, 
          omgevingsadressendichtheid, 
          stedelijkheid, 
          geom, 
          year)|>
  
  relocate(crs28992res500m, geom, year,
          aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland,
          percentage_laag_inkomen_huishouden,
          percentage_hoog_inkomen_huishouden, 
          omgevingsadressendichtheid, 
          stedelijkheid) |>
  
    rename(
          aantal_inwoners_500 = aantal_inwoners,
          aantal_inwoners_0_tot_15_jaar_500 = aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar_500 = aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar_500 = aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar_500 = aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder_500 = aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland_500 = percentage_geb_nederland_herkomst_nederland
  ) |>
  
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995)))
```

Now we join the squares with the postal code areas based on the largest overlapping area.

```{r eval =FALSE}
#match the squares belonging to the postal code area and keep only the square coordinates
df <- df |>
  st_join(cbs_500,join = st_intersects, left = T, largest = T) |>
  select(-c(year.y, aantal_inwoners_500, aantal_inwoners_0_tot_15_jaar_500,
          aantal_inwoners_15_tot_25_jaar_500,
          aantal_inwoners_25_tot_45_jaar_500,
          aantal_inwoners_45_tot_65_jaar_500,
          aantal_inwoners_65_jaar_en_ouder_500,
          percentage_geb_nederland_herkomst_nederland_500, percentage_laag_inkomen_huishouden, percentage_hoog_inkomen_huishouden, omgevingsadressendichtheid, stedelijkheid))

#match the square attributes in the right year
df <- df |>
  left_join (st_drop_geometry(cbs_500), by = c("crs28992res500m", "year.x" = "year")) |>
  rename(year = year.x) |>
  mutate(across(where(is.numeric), ~na_if(., NaN)))

```

Saving the data!

```{r eval=FALSE}
st_write(df, "./data/merged_data.gpkg", append = FALSE)

```
