---
title: "Journal 3"
author: "Kalle Stoffers"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=F)
```

## Exploring the data

Relevant packages

```{r}
rm(list=ls())

library(tidyverse)
library(sf)
library(tmap)
library(spdep)
library(TTR)

```

```{r}
df <- st_read(dsn = "./data/merged_data.gpkg")

```

Looking at the solar panel data again

```{r}
#leveringsrichting from character to numeric
class(df$leveringsrichting)
df$leveringsrichting <- as.numeric(gsub(",", ".", df$leveringsrichting))

#reverse 
df<- df|> mutate(teruglevering = 100 - leveringsrichting)
summary(df$teruglevering)
hist(df$teruglevering)
df %>% filter(teruglevering != 0) %>% nrow()

#dichotomous variable
df <- df |> mutate (teruglevering_d = if_else(teruglevering>0, 1, 0))
summary(df$teruglevering_d)

#plots over time
df |>
  group_by(year) |>
  summarise(yearly_avg = mean(teruglevering, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = yearly_avg)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(min(df$year), max(df$year), by = 2))+
  theme_minimal()

#dichotomous
df |>
  group_by(year) |>
  summarise(yearly_avg = mean(teruglevering_d, na.rm = TRUE) * 100) |>
  ggplot(aes(x = year, y = yearly_avg)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(min(df$year), max(df$year), by = 2))+
  theme_minimal()

```

Let's plot the data on a map using ggplot.

```{r}
#plot the data
df |>
  filter(year == 2024) |>
  ggplot(
    aes(fill = teruglevering)
  ) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "plasma", name = "Energy feed-in") +
  theme_minimal()



```

Now we have only mapped the data for a single year. Now, we create a map of all the years using tmap:

```{r}
tmap_mode("plot")

tm_shape(df) +
  tm_fill("teruglevering") +
  tm_facets(by = "year")

```

Calculating the installed base effect.

```{r}
#calculate the weighted sum of energy feed-in per 500m square per year.
avg_500 <- df |>
  st_drop_geometry() |>
  group_by(crs28992res500m, year) |>
  summarise(
    sum_teruglevering = sum((teruglevering * aantal_part_huishoudens), na.rm = TRUE),
    n_hh = sum(aantal_part_huishoudens),
    .groups = "drop"
  )

#select ego info at t-1
ego_t_minus_one <- df |>
  st_drop_geometry() |> 
  rename(teruglevering_t_minus_one = teruglevering,
         huishoudens_t_minus_one = aantal_part_huishoudens,
         crs28992res500m_t_minus_one = crs28992res500m) |> 
  select(postcode_van, year, crs28992res500m_t_minus_one, teruglevering_t_minus_one, huishoudens_t_minus_one)


#add the information at t-1 and making neighborhood average without "ego". 
df <- df |>
  mutate(year_t_minus_one = year - 1) |>
  
  left_join(ego_t_minus_one,
            by = join_by(postcode_van, year_t_minus_one == year)) |>
  #t-1
  left_join(avg_500,
            by = join_by(crs28992res500m_t_minus_one == crs28992res500m, year_t_minus_one == year)) |>
  #t0
  left_join(avg_500 |> rename(sum_teruglevering_t0 = sum_teruglevering, n_t0 = n_hh),
            by = join_by(crs28992res500m == crs28992res500m, year == year)) |>
  
  mutate(nb_pv = ifelse(
    is.na(huishoudens_t_minus_one),
    #use t0 if previous year not available
    (sum_teruglevering_t0 - (teruglevering * aantal_part_huishoudens)) / 
      (n_t0 - aantal_part_huishoudens),
    #use t-1 as default
    (sum_teruglevering - (teruglevering_t_minus_one * huishoudens_t_minus_one)) /
      (n_hh - huishoudens_t_minus_one))) 
  

#delete rows with zero residents
df <- df |> filter(aantal_inwoners != 0)

#descriptives
summary(df$teruglevering)
summary(df$nb_pv)

df |>
  filter(is.na(nb_pv)) |>
  st_drop_geometry() |>
  count(year)

#The remaining 1293 missing are the only streets in a square
df |>
  st_drop_geometry() |>
  filter(is.na(nb_pv)) |>
  mutate(alone_in_square = case_when(
    # default t-1 option
    !is.na(huishoudens_t_minus_one) ~ n_hh == huishoudens_t_minus_one,
    # when using t0 info
    is.na(huishoudens_t_minus_one) ~ n_t0 == aantal_part_huishoudens
  )) |>
  count(alone_in_square)

```

Make function to impute missing data in certain years based on the exponential moving average of that street

```{r}
df <- df |> arrange(postcode_van, year)

impute_ema <- function(df, target_year, history_years, var) {
  imputed <- df |>
    filter(is.na({{ var }}) & year == target_year) |>
    pull(postcode_van) |>
    unique() |>
    sapply(function(pc) {
      hist_values <- df |>
        filter(postcode_van == pc, year %in% history_years) |>
        arrange(year) |>
        pull({{ var }}) |>
        na.omit()
      
      if (length(hist_values) == 0 || all(is.na(hist_values))) {
        NA_real_
      } else if (length(hist_values) == 1) {
        hist_values[1]
      } else {
        tail(EMA(hist_values, 2), 1)
      }
    }) |>
    enframe(name = "postcode_van", value = "imputed_value")
  
  df |>
    left_join(imputed, by = "postcode_van") |>
    mutate({{ var }} := case_when(
      is.na({{ var }}) & year == target_year ~ imputed_value,
      .default = {{ var }}
    )) |>
    select(-imputed_value)
}

```

Impute missing years

```{r}

df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(percentage_hoog_inkomen_huishouden) & is.na(percentage_laag_inkomen_huishouden)))

df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddelde_woz_waarde_woning)))

df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddeld_elektriciteitsverbruik_woning)))


df <- df |>
  impute_ema(target_year = 2023, history_years = c(2021, 2022), var = percentage_laag_inkomen_huishouden) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = percentage_laag_inkomen_huishouden) |>
  impute_ema(target_year = 2023, history_years = c(2021, 2022), var = percentage_hoog_inkomen_huishouden) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = percentage_hoog_inkomen_huishouden) |>
  impute_ema(target_year = 2015, history_years = c(2017, 2016), var = gemiddelde_woz_waarde_woning) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = gemiddelde_woz_waarde_woning) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = gemiddeld_elektriciteitsverbruik_woning)



```

Calculating diversity measures: Herfindahl index.

```{r}
##economic diversity
df <- df |>
  mutate(
    economic_diversity = 1 - ((percentage_hoog_inkomen_huishouden^2 +
                                 percentage_laag_inkomen_huishouden^2 + 
                                 (100 - percentage_hoog_inkomen_huishouden - percentage_laag_inkomen_huishouden)^2
                                  ) 
         / 10000 
                                 )
         )

summary(df$economic_diversity)

##ethnic diversity 

df <- df |>
  mutate(
    ethnic_diversity = 1 - ((percentage_geb_nederland_herkomst_nederland_500^2 + (100 - percentage_geb_nederland_herkomst_nederland_500)^2) / 10000)
  )

summary(df$ethnic_diversity)

df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(ethnic_diversity)))

##age diversity

#set number in age category to 0 if the total number of people is not missing
df <- df |>
  mutate(
    aantal_inwoners_0_tot_15_jaar_500 = if_else(
      is.na(aantal_inwoners_0_tot_15_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_0_tot_15_jaar_500
    ),
    aantal_inwoners_15_tot_25_jaar_500 = if_else(
      is.na(aantal_inwoners_15_tot_25_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_15_tot_25_jaar_500
    ),
    aantal_inwoners_25_tot_45_jaar_500 = if_else(
      is.na(aantal_inwoners_25_tot_45_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_25_tot_45_jaar_500
    ),
    aantal_inwoners_45_tot_65_jaar_500 = if_else(
      is.na(aantal_inwoners_45_tot_65_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_45_tot_65_jaar_500
    ),
    aantal_inwoners_65_jaar_en_ouder_500 = if_else(
      is.na(aantal_inwoners_65_jaar_en_ouder_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_65_jaar_en_ouder_500
    )
  )

#calculate proportions
df <- df |>
  mutate(
    percentage_inwoners_0_tot_15_jaar_500 = aantal_inwoners_0_tot_15_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_15_tot_25_jaar_500 = aantal_inwoners_15_tot_25_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_25_tot_45_jaar_500 = aantal_inwoners_25_tot_45_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_45_tot_65_jaar_500 = aantal_inwoners_45_tot_65_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_65_jaar_en_ouder_500 = aantal_inwoners_65_jaar_en_ouder_500 / aantal_inwoners_500,
  )

df <- df |>
  mutate(
    age_diversity = 1 - (percentage_inwoners_0_tot_15_jaar_500^2 +
                    percentage_inwoners_15_tot_25_jaar_500^2 +
                    percentage_inwoners_25_tot_45_jaar_500^2 +
                    percentage_inwoners_45_tot_65_jaar_500^2 +
                    percentage_inwoners_65_jaar_en_ouder_500^2)
  )


summary(df$age_diversity)

df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(age_diversity)))

```

Add data on solar panel prices (source: <https://www.irena.org/Data/View-data-by-topic/Costs/Solar-costs>)

```{r}
df <- df |>
  mutate(solar_cost = case_when(
    year == 2015 ~ NA,
    year == 2016 ~ 0.1291,
    year == 2017 ~ 0.1430,
    year == 2018 ~ 0.1092,
    year == 2019 ~ 0.1040,
    year == 2020 ~ 0.1006,
    year == 2021 ~ 0.0836,
    year == 2022 ~ 0.0912,
    year == 2023 ~ 0.0699,
    year == 2024 ~ NA
  ))

df <- df |>
    impute_ema(target_year = 2015, history_years = c(2017, 2016), var = solar_cost) |>
    impute_ema(target_year = 2024, history_years = c(2022, 2023), var = solar_cost)
  
summary(df$solar_cost)
```

Looking at descriptives.

```{r}
## income
summary(df$percentage_laag_inkomen_huishouden)
hist(df$percentage_laag_inkomen_huishouden)

summary(df$percentage_hoog_inkomen_huishouden)
hist(df$percentage_hoog_inkomen_huishouden)

##property value
summary(df$gemiddelde_woz_waarde_woning)
hist(df$gemiddelde_woz_waarde_woning)

## density/urbanity
summary(df$omgevingsadressendichtheid)
hist(df$omgevingsadressendichtheid)

summary(df$stedelijkheid)
hist(df$stedelijkheid)

```

Calculate electricity costs based on electricity price according to CBS (<https://opendata.cbs.nl/#/CBS/nl/dataset/85666NED/table>)

```{r}
## electricity use
summary(df$gemiddeld_elektriciteitsverbruik_woning)
hist(df$gemiddeld_elektriciteitsverbruik_woning)

## add electricity price (CBS)
df <- df |>
  mutate(electricity_price = case_when(
    #2015
    gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2015 ~ -0.024,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2015 ~ 0.166,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2015 ~ 0.196,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2015 ~ 0.208,
    #2016
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2016 ~ -0.054,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2016 ~ 0.140,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2016 ~ 0.168,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2016 ~ 0.180,
    #2017
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2017 ~ -0.075,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2017 ~ 0.129,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2017 ~ 0.161,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2017 ~ 0.176,
    #2018
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2018 ~ -0.036,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2018 ~ 0.150,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2018 ~ 0.178,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2018 ~ 0.191,
    #2019
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2019 ~ 0.110,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2019 ~ 0.199,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2019 ~ 0.205,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2019 ~ 0.213,
    #2020
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2020 ~ -0.220,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2020 ~ 0.072,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2020 ~ 0.139,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2020 ~ 0.185,
    #2021
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2021 ~ -0.239,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2021 ~ 0.066,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2021 ~ 0.136,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2021 ~ 0.185,
    #2022
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2022 ~ -0.544,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2022 ~ -0.038,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2022 ~ 0.106,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2022 ~ 0.201,
    #2023
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2023 ~ 0.020,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2023 ~ 0.230,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2023 ~ 0.316,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2023 ~ 0.389,
    #2024
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2024 ~ -0.068,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2024 ~ 0.174,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2024 ~ 0.235,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2024 ~ 0.270,
  ))

summary(df$electricity_price)

## calculate electricity cost
df <- df |>
  mutate(electricity_cost = electricity_price * gemiddeld_elektriciteitsverbruik_woning)

summary(df$electricity_cost)


```

Control variables

```{r}
## home ownership
summary(df$percentage_koopwoningen)
summary(df$percentage_huurwoningen)

df |> filter(is.na(percentage_koopwoningen) & is.na(percentage_huurwoningen)) |> summarise(count = n())

#unnecessary missings
df <- df |>
  mutate(percentage_koopwoningen = if_else (
    is.na(percentage_koopwoningen) & !is.na(percentage_huurwoningen),
    100 - percentage_huurwoningen,
    percentage_koopwoningen
  ))

## ethnicity
summary(df$percentage_geb_nederland_herkomst_nederland)
hist(df$percentage_geb_nederland_herkomst_nederland)

## age
sum(is.na(df$aantal_inwoners_0_tot_15_jaar) & is.na(df$aantal_inwoners_15_tot_25_jaar) & is.na(df$aantal_inwoners_25_tot_45_jaar) & is.na(df$aantal_inwoners_45_tot_65_jaar) & is.na(df$aantal_inwoners_65_jaar_en_ouder))


#estimating the average age
df <- df |>
  mutate(
    age_avg = (
      (7.5 * aantal_inwoners_0_tot_15_jaar) + 
      (20 * aantal_inwoners_15_tot_25_jaar) + 
      (35 * aantal_inwoners_25_tot_45_jaar) +
      (55 * aantal_inwoners_45_tot_65_jaar) +
      (75 * aantal_inwoners_65_jaar_en_ouder)  
              ) 
    / aantal_inwoners
        )

summary(df$age_avg)
hist(df$age_avg)

```

Calculate the spatial autocorrelation. Appears to increase slightly over time (0.25 -\> 0.30).

```{r}
years <- c(2015:2018, 2021:2024)
moran_results_list <- list()

for (yr in years) {
  
  # Filter for the specific year
  df_yr <- df |> filter(year == yr)
    
    # Create neighbor list
    nb <- poly2nb(df_yr)
    
    # Create weights list
    lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
    
    # Run Moran's I test and store it in the list using the year as the name
    moran_results_list[[as.character(yr)]] <- moran.test(df_yr$teruglevering, 
                                                         lw, 
                                                         zero.policy = TRUE)
    
}

str(moran_results_list)
```

Clean columns a bit

```{r}

df <- df |>
  select (-c(year_t_minus_one, n_hh, n_t0, teruglevering_t_minus_one, huishoudens_t_minus_one, sum_teruglevering_t0, crs28992res500m_t_minus_one)) |>
  relocate()

```

Save data ready to use for analysis

```{r}
st_write(df, "./data/data_analysis.gpkg", append = FALSE)
```
