---
title: "Journal 3"
author: "Kalle Stoffers"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring the data

Relevant packages

```{r}
rm(list=ls())

library(tidyverse)
library(sf)
library(tmap)
library(spdep)

```

```{r}
df <- st_read(dsn = "./data/merged_data.gpkg")

```

Looking at the solar panel data again

```{r}
#leveringsrichting from character to numeric
class(df$leveringsrichting)
df$leveringsrichting <- as.numeric(gsub(",", ".", df$leveringsrichting))

#reverse 
df<- df|> mutate(teruglevering = 100 - leveringsrichting)
summary(df$teruglevering)
hist(df$teruglevering)
df %>% filter(teruglevering != 0) %>% nrow()

#dichotomous variable
df <- df |> mutate (teruglevering_d = if_else(teruglevering>0, 1, 0))
summary(df$teruglevering_d)

#plots over time
df |>
  group_by(year) |>
  summarise(yearly_avg = mean(teruglevering, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = yearly_avg)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(min(df$year), max(df$year), by = 2))+
  theme_minimal()

#dichotomous
df |>
  group_by(year) |>
  summarise(yearly_avg = mean(teruglevering_d, na.rm = TRUE) * 100) |>
  ggplot(aes(x = year, y = yearly_avg)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(min(df$year), max(df$year), by = 2))+
  theme_minimal()

```

Let's plot the data on a map using ggplot.

```{r}
#plot the data
df |>
  filter(year == 2024) |>
  ggplot(
    aes(fill = teruglevering)
  ) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "plasma", name = "Energy feed-in") +
  theme_minimal()



```

Now we have only mapped the data for a single year. Now, we create a map of all the years using tmap:

```{r}
tmap_mode("plot")

tm_shape(df) +
  tm_fill("teruglevering") +
  tm_facets(by = "year")

```

Looking at relevant variables. High levels of missingness in the data! Maybe I can use spatial + temporal interpolation to impute the data?

```{r}
## ethniticity
summary(df$percentage_geb_nederland_herkomst_nederland) #565 NAs
sum(
    is.na(df$percentage_geb_nederland_herkomst_nederland) & 
    is.na(df$percentage_geb_nederland_herkomst_overig_europa) & 
    is.na(df$percentage_geb_nederland_herkomst_buiten_europa) & 
    is.na(df$percentage_geb_buiten_nederland_herkomst_europa) &
    is.na(df$percentage_geb_buiten_nederland_herkmst_buiten_europa) &
    is.na(df$percentage_westerse_migr_achtergr) &
    is.na(df$percentage_niet_westerse_migr_achtergr)  
    ) #533 truly missing, 32 likely value close to 0. 

hist(df$percentage_geb_nederland_herkomst_nederland)

## home ownership
summary(df$percentage_koopwoningen) #4836 NAs
df |> filter(is.na(percentage_koopwoningen) & is.na(percentage_huurwoningen)) |> summarise(count = n()) #3958 of those have missing value on both rented and owned homes
df |> filter(is.na(percentage_koopwoningen) & !is.na(percentage_huurwoningen)) |> summarise(count = n()) #878 streets likely have close to 0 home ownership

df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(percentage_koopwoningen))) # about 600 NAs in every year

hist(df$percentage_koopwoningen)

## property value
summary(df$gemiddelde_woz_waarde_woning) #6219 NAs
df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddelde_woz_waarde_woning))) #no data in 2015 and 2024

hist(df$gemiddelde_woz_waarde_woning)

## electricity use
summary(df$gemiddeld_elektriciteitsverbruik_woning) #3476 NAs
df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddeld_elektriciteitsverbruik_woning))) #no data in 2024

hist(df$gemiddeld_elektriciteitsverbruik_woning)

## age
sum(is.na(df$aantal_inwoners_0_tot_15_jaar) & is.na(df$aantal_inwoners_15_tot_25_jaar) & is.na(df$aantal_inwoners_25_tot_45_jaar) & is.na(df$aantal_inwoners_45_tot_65_jaar) & is.na(df$aantal_inwoners_65_jaar_en_ouder)) #0 NAs!

sum(df$aantal_inwoners == 0) #but 411 streets with 0 residents

#estimating the average age
df <- df |>
  mutate(
    age_avg = (
      (7.5 * aantal_inwoners_0_tot_15_jaar) + 
      (20 * aantal_inwoners_15_tot_25_jaar) + 
      (35 * aantal_inwoners_25_tot_45_jaar) +
      (55 * aantal_inwoners_45_tot_65_jaar) +
      (75 * aantal_inwoners_65_jaar_en_ouder)  
              ) 
    / aantal_inwoners
        )

summary(df$age_avg)
hist(df$age_avg)

## income
summary(df$gemiddeld_inkomen_huishouden) #15858 NAs
df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddeld_inkomen_huishouden))) #data only from 2021 and 2022

hist(df$gemiddeld_inkomen_huishouden)

summary(df$percentage_laag_inkomen_huishouden) #7793 NAs
df |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(percentage_laag_inkomen_huishouden))) #no data in 2023 and 2024

hist(df$percentage_laag_inkomen_huishouden)

summary(df$percentage_hoog_inkomen_huishouden) #7796 NAs
hist(df$percentage_hoog_inkomen_huishouden)


## density/urbanity
summary(df$omgevingsadressendichtheid) #106 NAs
hist(df$omgevingsadressendichtheid)

summary(df$stedelijkheid) #106 NAs
hist(df$stedelijkheid)

```

Calculate the spatial autocorrelation.

```{r}

df_24 <- df |> filter (year == 2024)

nb <- poly2nb(df_24)

lw <- nb2listw(nb, style = "W", zero.policy = T)

moran_result <- moran.test(df_24$teruglevering, lw, zero.policy = TRUE)
```

Analysis:

-   Spatial lag panel model (slpm package)?

-   Social network analysis?

-   Panel/multilevel regression with installed base effect (energy feed-in by "neighbors" in the previous year)

    -   This allows you to interact on the installed-base variable. But:

        -   For the effect of density/urbanity, you have focus on the influence between two streets in the same 500m by 500m square. How?

        -   For the effect of demographic similarity, you have to calculate the similarity for each combination of streets and use that in the interaction with the installed-base variable. How?

-   Weight matrix by distance or "queen" contiguity?

-   For the effect of lucrativity, look at period effect? Or add data per year with installation price and eletricity price?
