---
title: "Preparing_data_clean"
author: "Kalle Stoffers"
date: "2026-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r}
rm(list=ls())

library(tidyverse)
library(sf)
library(spdep)
library(TTR)
```

## Step 1: PV data

To measure the amount of solar panels, we use open data from energy provider Westland Infra (<https://westlandinfra.nl/over-westland-infra/duurzaamheid-innovaties/open-data/>). As a dependent variable, we use the percentage of connections ("aansluitingen") in a street that have a net negative electricity usage, as this is indicative of the solar panel adoption in the street.

```{r}
pv <- readRDS("./data/pv_data.Rds") |>
  filter(year >= 2015) |>
  mutate(id = row_number()) |>
  relocate(id) |>
  rename(postcode_van = POSTCODE_VAN,
         postcode_tot = POSTCODE_TOT,
         straatnaam = STRAATNAAM,
         leveringsrichting = X.Leveringsrichting) |>
  select(id, postcode_van, postcode_tot, straatnaam, year, leveringsrichting)

class(pv$leveringsrichting)
pv$leveringsrichting <- as.numeric(gsub(",", ".", pv$leveringsrichting))

#the original variable represents the percentage with a positive electricity usage, we are interested in the opposite so we reverse the variable
pv <- pv |> mutate(teruglevering = 100 - leveringsrichting)

```

## Step 2: PC6 data

We can connect these streets in the Westland region to demographic data at the postal code level from the CBS. These data can be downloaded here: <https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/gegevens-per-postcode>. After 1st of may the data of 2019 and 2020 should be available as well!

```{r}
#load data
cbs_pc6_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_pc6_23 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_pc6_22 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_pc6_21 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2021_vol.gpkg") |>
  mutate (year = 2021) |>
  rename (percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)

cbs_pc6_2124 <- bind_rows(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22, cbs_pc6_21) |>
  select (- c(10, 17:20, 23:31, 37, 40:131))

rm(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22, cbs_pc6_21)
gc()

cbs_pc6_18 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2018-v3/CBS_PC6_2018_v3.shp") |>  
  mutate (year = 2018) |>
  rename (Postcode = PC6)
cbs_pc6_17 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2017-v3/CBS_PC6_2017_v3.shp") |>
  mutate (year = 2017)
cbs_pc6_16 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2016-v2/CBS_PC6_2016_v2.shp") |>
  mutate (year = 2016)
cbs_pc6_15 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2015-v2/CBS_PC6_2015_v2.shp") |>
  mutate (year = 2015)

cbs_pc6_1518 <- bind_rows(cbs_pc6_18, cbs_pc6_17, cbs_pc6_16, cbs_pc6_15) |>
  rename (
    postcode6 = Postcode,
    aantal_inwoners = INWONER,
    aantal_mannen = MAN,
    aantal_vrouwen = VROUW,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    aantal_part_huishoudens = AANTAL_HH,
    gemiddelde_huishoudensgrootte = GEM_HH_GR,
    aantal_woningen = WONING,
    percentage_koopwoningen = P_KOOPWON, 
    percentage_huurwoningen = P_HUURWON,
    aantal_huurwoningen_in_bezit_woningcorporaties = WON_HCORP,
    aantal_niet_bewoonde_woningen = WON_NBEW,
    gemiddelde_woz_waarde_woning = WOZWONING, 
    gemiddeld_elektriciteitsverbruik_woning = G_ELEK_WON,
    aantal_personen_met_uitkering_onder_aowlft = UITKMINAOW,
    geom = geometry
  ) |>
  select (- c(10, 15:18, 21:29, 35, 39:130))
  
rm(cbs_pc6_18, cbs_pc6_17, cbs_pc6_16, cbs_pc6_15)
gc()

#join and set NAs
cbs_pc6 <- bind_rows(cbs_pc6_2124, cbs_pc6_1518) |>
  relocate (postcode6, year, geom)|>
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995))) 

rm(cbs_pc6_2124, cbs_pc6_1518)
gc()
```

Some streets in the PV data span across multiple PC6 postal codes. Therefore we first select all of the postal codes of the CBS data that are included in the PV data.

```{r}
cbs_pc6_westland <- cbs_pc6 |>
  inner_join(
    pv, 
    by = join_by(
      year == year,
      postcode6 >= postcode_van,
      postcode6 <= postcode_tot
    )
  ) 

rm(cbs_pc6)
gc()
```

Now we have to transform the data from the single postal codes of the CBS to the areas containing one or more postal codes in the PV data.

```{r}
df_pc6 <- cbs_pc6_westland |>
  group_by(id) |>
  summarise(
    
    geometry = sf::st_union(geom),
    
    across(
      c(
        postcode_van, 
        postcode_tot, 
        straatnaam, 
        year,
        teruglevering
        ), 
           first),
    
    across(
      c(
        percentage_geb_nederland_herkomst_nederland,
        percentage_geb_nederland_herkomst_overig_europa,
        percentage_geb_nederland_herkomst_buiten_europa,
        percentage_geb_buiten_nederland_herkomst_europa,
        percentage_geb_buiten_nederland_herkmst_buiten_europa,
        percentage_westerse_migr_achtergr,
        percentage_niet_westerse_migr_achtergr
        ),
      ~ weighted.mean(.x, aantal_inwoners, na.rm = TRUE)
    ),
    
    gemiddelde_huishoudensgrootte = 
      weighted.mean(gemiddelde_huishoudensgrootte,
                    aantal_part_huishoudens, 
                    na.rm = TRUE),
    
    across(
      c(
        percentage_koopwoningen, 
        percentage_huurwoningen, 
        gemiddelde_woz_waarde_woning, 
        gemiddeld_elektriciteitsverbruik_woning
        ), 
           ~weighted.mean(.x, aantal_woningen, na.rm = TRUE)
      ),
    
    across(
      c(
        aantal_inwoners,
        aantal_mannen,
        aantal_vrouwen,
        aantal_inwoners_0_tot_15_jaar,
        aantal_inwoners_15_tot_25_jaar,
        aantal_inwoners_25_tot_45_jaar,
        aantal_inwoners_45_tot_65_jaar,
        aantal_inwoners_65_jaar_en_ouder,
        aantal_part_huishoudens
      ),
      ~ sum(.x, na.rm = TRUE)
      )
  )

rm(cbs_pc6_westland)
gc()
```

## Step 3: 500m \* 500m data

Some data, such as density and income is only available at the level of 500m by 500m squares. We also use these squares to calculate the demographic diversity of the surrounding neighborhood. These will also be the areas used to calculate the effect of neighborhood solar panels. The data can be downloaded here: [https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/kaart-van-500-meter-bij-500-meter-met-statistieken](#0){.uri}

```{r}
cbs_500_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_500_23 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2023_v2/cbs_vk500_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_500_22 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2022_vol/cbs_vk500_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_500_21 <- st_read(dsn = "C:/Users/kalle/Downloads/2024-cbs_vk500_2021_vol/cbs_vk500_2021_vol.gpkg") |>
  mutate (year = 2021) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_20 <- st_read(dsn = "C:/Users/kalle/Downloads/2023-cbs_vk500_2020_vol/cbs_vk500_2020_vol.gpkg") |>
  mutate (year = 2020) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_19 <- st_read(dsn = "C:/Users/kalle/Downloads/2023-cbs_vk500_2019_vol2/cbs_vk500_2019_vol2.gpkg") |>
  mutate (year = 2019) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_18 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2018.gpkg") |>
  mutate (year = 2018) |>
  rename(crs28992res500m = c28992r500, 
         percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_17 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2017.gpkg") |>
  mutate (year = 2017) |>
  rename(crs28992res500m = c28992r500,
         percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)

cbs_500_16 <- st_read(dsn = "C:/Users/kalle/Downloads/2019-cbs_vk500_2016_v2/CBS_VK500_2016_v2.shp") |>
  mutate (year = 2016) |>
  rename(
    crs28992res500m = C28992R500,
    aantal_inwoners = INWONER,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    percentage_laag_inkomen_huishouden = P_LINK_HH, 
    percentage_hoog_inkomen_huishouden = P_HINK_HH, 
    omgevingsadressendichtheid = OAD, 
    stedelijkheid = STED, 
    geom = geometry
  )

cbs_500_15 <- st_read(dsn = "C:/Users/kalle/Downloads/2019-cbs_vk500_2015_v2/CBS_VK500_2015_v2.shp") |>
  mutate (year = 2015) |>
  rename(
    crs28992res500m = C28992R500,
    aantal_inwoners = INWONER,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    percentage_laag_inkomen_huishouden = P_LINK_HH, 
    percentage_hoog_inkomen_huishouden = P_HINK_HH, 
    omgevingsadressendichtheid = OAD, 
    stedelijkheid = STED, 
    geom = geometry
  )

cbs_500 <- bind_rows(cbs_500_24, cbs_500_23, cbs_500_22, cbs_500_21, cbs_500_20, cbs_500_19, cbs_500_18, cbs_500_17, cbs_500_16, cbs_500_15) 

rm(cbs_500_24, cbs_500_23, cbs_500_22, cbs_500_21, cbs_500_20, cbs_500_19, cbs_500_18, cbs_500_17, cbs_500_16, cbs_500_15)
gc()

cbs_500 <- cbs_500 |>
  
  select(crs28992res500m, 
          aantal_inwoners,
          aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland,
          percentage_laag_inkomen_huishouden,
          percentage_hoog_inkomen_huishouden, 
          omgevingsadressendichtheid, 
          stedelijkheid, 
          geom, 
          year)|>
  
  relocate(crs28992res500m, geom, year,
          aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland,
          percentage_laag_inkomen_huishouden,
          percentage_hoog_inkomen_huishouden, 
          omgevingsadressendichtheid, 
          stedelijkheid) |>
  
    rename(
          aantal_inwoners_500 = aantal_inwoners,
          aantal_inwoners_0_tot_15_jaar_500 = aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar_500 = aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar_500 = aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar_500 = aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder_500 = aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland_500 = percentage_geb_nederland_herkomst_nederland
  ) |>
  
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995)))

```

Add the square data to the pc6 data

```{r}
#match the squares belonging to the postal code area and keep only the square coordinates
df_cbs <- df_pc6 |>
  st_join(cbs_500,join = st_intersects, left = T, largest = T) |>
  select(-c(year.y, aantal_inwoners_500, aantal_inwoners_0_tot_15_jaar_500,
          aantal_inwoners_15_tot_25_jaar_500,
          aantal_inwoners_25_tot_45_jaar_500,
          aantal_inwoners_45_tot_65_jaar_500,
          aantal_inwoners_65_jaar_en_ouder_500,
          percentage_geb_nederland_herkomst_nederland_500, percentage_laag_inkomen_huishouden, percentage_hoog_inkomen_huishouden, omgevingsadressendichtheid, stedelijkheid))

#match the square attributes in the right year
df_cbs <- df_cbs |>
  left_join (st_drop_geometry(cbs_500), by = c("crs28992res500m", "year.x" = "year")) |>
  rename(year = year.x) |>
  mutate(across(where(is.numeric), ~na_if(., NaN)))
```

Save the raw CBS data

```{r}
st_write(df_cbs, "./data/raw_df_cbs.gpkg", append = FALSE)
rm(cbs_500, df_pc6)
```

## Step 4: calculate relevant variables and impute missing years

```{r}
df_cbs <- st_read("./data/raw_df_cbs.gpkg")
```

Some variables in the dataset are completely missing in some years. So we can use a streets' previous values to predict the value in the missing years using the exponential moving average. To do this for different years and variables, we write a function.

```{r}

df_cbs <- df_cbs |> arrange(postcode_van, year)

impute_ema <- function(df, target_year, history_years, var) {
  imputed <- df |>
    filter(is.na({{ var }}) & year == target_year) |>
    pull(postcode_van) |>
    unique() |>
    sapply(function(pc) {
      hist_values <- df |>
        filter(postcode_van == pc, year %in% history_years) |>
        arrange(year) |>
        pull({{ var }}) |>
        na.omit()
      
      if (length(hist_values) == 0 || all(is.na(hist_values))) {
        NA_real_
      } else if (length(hist_values) == 1) {
        hist_values[1]
      } else {
        tail(EMA(hist_values, 2), 1)
      }
    }) |>
    enframe(name = "postcode_van", value = "imputed_value")
  
  df |>
    left_join(imputed, by = "postcode_van") |>
    mutate({{ var }} := case_when(
      is.na({{ var }}) & year == target_year ~ imputed_value,
      .default = {{ var }}
    )) |>
    select(-imputed_value)
}
```

Apply function.

```{r}
#delete rows with zero residents
df_cbs <- df_cbs |> filter(aantal_inwoners != 0)

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(percentage_hoog_inkomen_huishouden) & is.na(percentage_laag_inkomen_huishouden)))

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddelde_woz_waarde_woning)))

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddeld_elektriciteitsverbruik_woning)))


df_cbs <- df_cbs |>
  impute_ema(target_year = 2023, history_years = c(2021, 2022), var = percentage_laag_inkomen_huishouden) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = percentage_laag_inkomen_huishouden) |>
  impute_ema(target_year = 2023, history_years = c(2021, 2022), var = percentage_hoog_inkomen_huishouden) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = percentage_hoog_inkomen_huishouden) |>
  impute_ema(target_year = 2015, history_years = c(2017, 2016), var = gemiddelde_woz_waarde_woning) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = gemiddelde_woz_waarde_woning) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = gemiddeld_elektriciteitsverbruik_woning)

```

Calculating demographic diversity in the 500m x 500m squares using the Herfindahl index.

```{r}
##economic diversity
df_cbs <- df_cbs |>
  mutate(
    economic_diversity = 1 - ((percentage_hoog_inkomen_huishouden^2 +
                                 percentage_laag_inkomen_huishouden^2 + 
                                 (100 - percentage_hoog_inkomen_huishouden - percentage_laag_inkomen_huishouden)^2
                                  ) 
         / 10000 
                                 )
         )

summary(df_cbs$economic_diversity)

##ethnic diversity 

df_cbs <- df_cbs |>
  mutate(
    ethnic_diversity = 1 - ((percentage_geb_nederland_herkomst_nederland_500^2 + (100 - percentage_geb_nederland_herkomst_nederland_500)^2) / 10000)
  )

summary(df_cbs$ethnic_diversity)

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(ethnic_diversity)))

##age diversity

#set number in age category to 0 if the total number of people is not missing
#this number is set to missing if there are less than 4 people in the category
#while we actually do have information on it
df_cbs <- df_cbs |>
  mutate(
    aantal_inwoners_0_tot_15_jaar_500 = if_else(
      is.na(aantal_inwoners_0_tot_15_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_0_tot_15_jaar_500
    ),
    aantal_inwoners_15_tot_25_jaar_500 = if_else(
      is.na(aantal_inwoners_15_tot_25_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_15_tot_25_jaar_500
    ),
    aantal_inwoners_25_tot_45_jaar_500 = if_else(
      is.na(aantal_inwoners_25_tot_45_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_25_tot_45_jaar_500
    ),
    aantal_inwoners_45_tot_65_jaar_500 = if_else(
      is.na(aantal_inwoners_45_tot_65_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_45_tot_65_jaar_500
    ),
    aantal_inwoners_65_jaar_en_ouder_500 = if_else(
      is.na(aantal_inwoners_65_jaar_en_ouder_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_65_jaar_en_ouder_500
    )
  )

#calculate proportions
df_cbs <- df_cbs |>
  mutate(
    percentage_inwoners_0_tot_15_jaar_500 = aantal_inwoners_0_tot_15_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_15_tot_25_jaar_500 = aantal_inwoners_15_tot_25_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_25_tot_45_jaar_500 = aantal_inwoners_25_tot_45_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_45_tot_65_jaar_500 = aantal_inwoners_45_tot_65_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_65_jaar_en_ouder_500 = aantal_inwoners_65_jaar_en_ouder_500 / aantal_inwoners_500,
  )

#calculate diversity index
df_cbs <- df_cbs |>
  mutate(
    age_diversity = 1 - (percentage_inwoners_0_tot_15_jaar_500^2 +
                    percentage_inwoners_15_tot_25_jaar_500^2 +
                    percentage_inwoners_25_tot_45_jaar_500^2 +
                    percentage_inwoners_45_tot_65_jaar_500^2 +
                    percentage_inwoners_65_jaar_en_ouder_500^2)
  )


summary(df_cbs$age_diversity)

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(age_diversity)))

```

Add historic solar panel costs from IRENA (<https://www.irena.org/Data/View-data-by-topic/Costs/Solar-costs>). This is expressed in USD per kWh and represents the "levelized cost of electricity" (lcoe), which are all the costs during the expected lifetime of the solar panel, divided by all the electricity produced during the same period.

```{r}
df_cbs <- df_cbs |>
  mutate(solar_cost = case_when(
    year == 2015 ~ NA,
    year == 2016 ~ 0.1291,
    year == 2017 ~ 0.1430,
    year == 2018 ~ 0.1092,
    year == 2019 ~ 0.1040,
    year == 2020 ~ 0.1006,
    year == 2021 ~ 0.0836,
    year == 2022 ~ 0.0912,
    year == 2023 ~ 0.0699,
    year == 2024 ~ NA
  ))

df_cbs <- df_cbs |>
    impute_ema(target_year = 2015, history_years = c(2017, 2016), var = solar_cost) |>
    impute_ema(target_year = 2024, history_years = c(2022, 2023), var = solar_cost)
  
summary(df_cbs$solar_cost)
```

Add electricity price data from CBS (<https://opendata.cbs.nl/#/CBS/nl/dataset/85666NED/table>). The average electricity price is different for different usage quantities.

```{r}
## electricity use
summary(df_cbs$gemiddeld_elektriciteitsverbruik_woning)
hist(df_cbs$gemiddeld_elektriciteitsverbruik_woning)

## add electricity price (CBS)
df_cbs <- df_cbs |>
  mutate(electricity_price = case_when(
    #2015
    gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2015 ~ -0.024,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2015 ~ 0.166,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2015 ~ 0.196,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2015 ~ 0.208,
    #2016
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2016 ~ -0.054,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2016 ~ 0.140,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2016 ~ 0.168,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2016 ~ 0.180,
    #2017
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2017 ~ -0.075,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2017 ~ 0.129,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2017 ~ 0.161,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2017 ~ 0.176,
    #2018
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2018 ~ -0.036,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2018 ~ 0.150,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2018 ~ 0.178,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2018 ~ 0.191,
    #2019
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2019 ~ 0.110,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2019 ~ 0.199,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2019 ~ 0.205,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2019 ~ 0.213,
    #2020
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2020 ~ -0.220,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2020 ~ 0.072,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2020 ~ 0.139,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2020 ~ 0.185,
    #2021
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2021 ~ -0.239,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2021 ~ 0.066,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2021 ~ 0.136,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2021 ~ 0.185,
    #2022
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2022 ~ -0.544,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2022 ~ -0.038,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2022 ~ 0.106,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2022 ~ 0.201,
    #2023
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2023 ~ 0.020,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2023 ~ 0.230,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2023 ~ 0.316,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2023 ~ 0.389,
    #2024
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2024 ~ -0.068,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2024 ~ 0.174,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2024 ~ 0.235,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2024 ~ 0.270,
  ))

summary(df_cbs$electricity_price)

## calculate electricity cost
df_cbs <- df_cbs |>
  mutate(electricity_cost = electricity_price * gemiddeld_elektriciteitsverbruik_woning)

summary(df_cbs$electricity_cost)
```

Look at control variables and their missing values

```{r}
## home ownership
summary(df_cbs$percentage_koopwoningen)
summary(df_cbs$percentage_huurwoningen)

df_cbs |> filter(is.na(percentage_koopwoningen) & is.na(percentage_huurwoningen)) |> summarise(count = n())

#unnecessary missings
#if we know either the percentage of rental homes or the percentage of owned homes
#we can calculate the other percentage
df_cbs <- df_cbs |>
  mutate(percentage_koopwoningen = if_else (
    is.na(percentage_koopwoningen) & !is.na(percentage_huurwoningen),
    100 - percentage_huurwoningen,
    percentage_koopwoningen
  ))

## ethnicity
summary(df_cbs$percentage_geb_nederland_herkomst_nederland)
hist(df_cbs$percentage_geb_nederland_herkomst_nederland)

## age


#estimating the average age from the categories
df_cbs <- df_cbs |>
  mutate(
    age_avg = (
      (7.5 * aantal_inwoners_0_tot_15_jaar) + 
      (20 * aantal_inwoners_15_tot_25_jaar) + 
      (35 * aantal_inwoners_25_tot_45_jaar) +
      (55 * aantal_inwoners_45_tot_65_jaar) +
      (75 * aantal_inwoners_65_jaar_en_ouder)  
              ) 
    / aantal_inwoners
        )

summary(df_cbs$age_avg)
hist(df_cbs$age_avg)
```

### Installed base

Calculating the average energy feed-in in the same 500m x 500m square (weighted by number of households), to represent the effect of social influence by nearby solar panels. I use the 500x500 square as neighbors because many of the moderators are most easily available at this level.

```{r}

#calculate the weighted sum of energy feed-in per 500m square per year.
avg_500 <- df_cbs |>
  st_drop_geometry() |>
  group_by(crs28992res500m, year) |>
  summarise(
    nb_pv = sum((teruglevering * aantal_part_huishoudens) / sum(aantal_part_huishoudens), na.rm = T),
    .groups = "drop")

#calculate
df_cbs <- df_cbs |> 
  left_join(avg_500, by = c("crs28992res500m", "year"))

summary(df_cbs$teruglevering)
summary(df_cbs$nb_pv)

```

## Step 5: Join based on t-1

```{r}
#only select the colums I may use
df_cbs <- df_cbs |>
  select(-c(teruglevering, percentage_geb_nederland_herkomst_overig_europa, percentage_geb_nederland_herkomst_buiten_europa, percentage_geb_buiten_nederland_herkomst_europa, percentage_geb_buiten_nederland_herkmst_buiten_europa, percentage_niet_westerse_migr_achtergr, percentage_westerse_migr_achtergr, gemiddelde_huishoudensgrootte, aantal_mannen, aantal_vrouwen, aantal_inwoners_0_tot_15_jaar_500, aantal_inwoners_15_tot_25_jaar_500, aantal_inwoners_25_tot_45_jaar_500, aantal_inwoners_45_tot_65_jaar_500, aantal_inwoners_65_jaar_en_ouder_500, percentage_inwoners_0_tot_15_jaar_500, percentage_inwoners_15_tot_25_jaar_500, percentage_inwoners_25_tot_45_jaar_500, percentage_inwoners_45_tot_65_jaar_500, percentage_inwoners_65_jaar_en_ouder_500))
  
```

Match the demographic predictor data to the solar panel data a year later, given the delay between the decision to adopt solar panels and the installation. Therefore the causal claim is stronger.

```{r}
#join at t0 if t+1 is not available??

df <- pv |>
  mutate(year_t_minus_one = year - 1) |>
  inner_join(
    df_cbs,
    by = join_by(postcode_van == postcode_van,
                 year_t_minus_one == year)
  ) |>
  rename(id = id.x,
         postcode_tot = postcode_tot.x,
         straatnaam = straatnaam.x) |>
  select( -c(id.y, postcode_tot.y, straatnaam.y, year_t_minus_one)) |>
  relocate(id, postcode_van, postcode_tot, straatnaam, year, teruglevering, nb_pv, percentage_hoog_inkomen_huishouden, percentage_laag_inkomen_huishouden, gemiddelde_woz_waarde_woning, solar_cost, gemiddeld_elektriciteitsverbruik_woning, electricity_price, electricity_cost, economic_diversity, ethnic_diversity, age_diversity, omgevingsadressendichtheid, stedelijkheid, percentage_geb_nederland_herkomst_nederland, age_avg, percentage_koopwoningen, percentage_huurwoningen)
```

Save data!

```{r}

st_write(df, "./data/df_analysis.gpkg", append = F)

```
