---
title: "Preparing_data_clean"
author: "Kalle Stoffers"
date: "2026-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r}
rm(list=ls())

library(tidyverse)
library(sf)
library(spdep)
library(TTR)
```

## Step 1: PV data

```{r}
pv <- readRDS("./data/pv_data.Rds") |>
  filter(year >= 2015) |>
  mutate(id = row_number()) |>
  relocate(id) |>
  rename(postcode_van = POSTCODE_VAN,
         postcode_tot = POSTCODE_TOT,
         straatnaam = STRAATNAAM,
         leveringsrichting = X.Leveringsrichting) |>
  select(id, postcode_van, postcode_tot, straatnaam, year, leveringsrichting)

class(pv$leveringsrichting)
pv$leveringsrichting <- as.numeric(gsub(",", ".", pv$leveringsrichting))

#reverse 
pv <- pv |> mutate(teruglevering = 100 - leveringsrichting)

```

## Step 2: PC6 data

These data can be downloaded here: <https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/gegevens-per-postcode>. After 1st of may the data of 2019 and 2020 should be available as well!

```{r}
#load data
cbs_pc6_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_pc6_23 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_pc6_22 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_pc6_21 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_pc6_2021_vol.gpkg") |>
  mutate (year = 2021) |>
  rename (percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)

cbs_pc6_2124 <- bind_rows(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22, cbs_pc6_21) |>
  select (- c(10, 17:20, 23:31, 37, 40:131))

rm(cbs_pc6_24, cbs_pc6_23, cbs_pc6_22, cbs_pc6_21)
gc()

cbs_pc6_18 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2018-v3/CBS_PC6_2018_v3.shp") |>  
  mutate (year = 2018) |>
  rename (Postcode = PC6)
cbs_pc6_17 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2017-v3/CBS_PC6_2017_v3.shp") |>
  mutate (year = 2017)
cbs_pc6_16 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2016-v2/CBS_PC6_2016_v2.shp") |>
  mutate (year = 2016)
cbs_pc6_15 <- st_read(dsn = "C:/Users/kalle/Downloads/CBS-PC6-2015-v2/CBS_PC6_2015_v2.shp") |>
  mutate (year = 2015)

cbs_pc6_1518 <- bind_rows(cbs_pc6_18, cbs_pc6_17, cbs_pc6_16, cbs_pc6_15) |>
  rename (
    postcode6 = Postcode,
    aantal_inwoners = INWONER,
    aantal_mannen = MAN,
    aantal_vrouwen = VROUW,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    aantal_part_huishoudens = AANTAL_HH,
    gemiddelde_huishoudensgrootte = GEM_HH_GR,
    aantal_woningen = WONING,
    percentage_koopwoningen = P_KOOPWON, 
    percentage_huurwoningen = P_HUURWON,
    aantal_huurwoningen_in_bezit_woningcorporaties = WON_HCORP,
    aantal_niet_bewoonde_woningen = WON_NBEW,
    gemiddelde_woz_waarde_woning = WOZWONING, 
    gemiddeld_elektriciteitsverbruik_woning = G_ELEK_WON,
    aantal_personen_met_uitkering_onder_aowlft = UITKMINAOW,
    geom = geometry
  ) |>
  select (- c(10, 15:18, 21:29, 35, 39:130))
  
rm(cbs_pc6_18, cbs_pc6_17, cbs_pc6_16, cbs_pc6_15)
gc()

#join and set NAs
cbs_pc6 <- bind_rows(cbs_pc6_2124, cbs_pc6_1518) |>
  relocate (postcode6, year, geom)|>
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995))) 

rm(cbs_pc6_2124, cbs_pc6_1518)
gc()
```

Select the postal codes of the CBS data that are included in the PV data

```{r}
cbs_pc6_westland <- cbs_pc6 |>
  inner_join(
    pv, 
    by = join_by(
      year == year,
      postcode6 >= postcode_van,
      postcode6 <= postcode_tot
    )
  ) 

rm(cbs_pc6)
gc()
```

Some streets in the pv data match with multiple pc6 postal codes, so we have to summarise them back together.

```{r}
df_pc6 <- cbs_pc6_westland |>
  group_by(id) |>
  summarise(
    
    geometry = sf::st_union(geom),
    
    across(
      c(
        postcode_van, 
        postcode_tot, 
        straatnaam, 
        year,
        teruglevering
        ), 
           first),
    
    across(
      c(
        percentage_geb_nederland_herkomst_nederland,
        percentage_geb_nederland_herkomst_overig_europa,
        percentage_geb_nederland_herkomst_buiten_europa,
        percentage_geb_buiten_nederland_herkomst_europa,
        percentage_geb_buiten_nederland_herkmst_buiten_europa,
        percentage_westerse_migr_achtergr,
        percentage_niet_westerse_migr_achtergr
        ),
      ~ weighted.mean(.x, aantal_inwoners, na.rm = TRUE)
    ),
    
    gemiddelde_huishoudensgrootte = 
      weighted.mean(gemiddelde_huishoudensgrootte,
                    aantal_part_huishoudens, 
                    na.rm = TRUE),
    
    across(
      c(
        percentage_koopwoningen, 
        percentage_huurwoningen, 
        gemiddelde_woz_waarde_woning, 
        gemiddeld_elektriciteitsverbruik_woning
        ), 
           ~weighted.mean(.x, aantal_woningen, na.rm = TRUE)
      ),
    
    across(
      c(
        aantal_inwoners,
        aantal_mannen,
        aantal_vrouwen,
        aantal_inwoners_0_tot_15_jaar,
        aantal_inwoners_15_tot_25_jaar,
        aantal_inwoners_25_tot_45_jaar,
        aantal_inwoners_45_tot_65_jaar,
        aantal_inwoners_65_jaar_en_ouder,
        aantal_part_huishoudens
      ),
      ~ sum(.x, na.rm = TRUE)
      )
  )

rm(cbs_pc6_westland)
gc()
```

## Step 3: 500m \* 500m data

This data can be downloaded here: [https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data/kaart-van-500-meter-bij-500-meter-met-statistieken](#0){.uri}

```{r}
cbs_500_24 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2024_v1.gpkg") |>
  mutate (year = 2024)
cbs_500_23 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2023_v2/cbs_vk500_2023_v2.gpkg") |>
  mutate (year = 2023)
cbs_500_22 <- st_read(dsn = "C:/Users/kalle/Downloads/2025-cbs_vk500_2022_vol/cbs_vk500_2022_vol.gpkg") |>
  mutate (year = 2022)
cbs_500_21 <- st_read(dsn = "C:/Users/kalle/Downloads/2024-cbs_vk500_2021_vol/cbs_vk500_2021_vol.gpkg") |>
  mutate (year = 2021) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_20 <- st_read(dsn = "C:/Users/kalle/Downloads/2023-cbs_vk500_2020_vol/cbs_vk500_2020_vol.gpkg") |>
  mutate (year = 2020) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_19 <- st_read(dsn = "C:/Users/kalle/Downloads/2023-cbs_vk500_2019_vol2/cbs_vk500_2019_vol2.gpkg") |>
  mutate (year = 2019) |>
  rename(percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_18 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2018.gpkg") |>
  mutate (year = 2018) |>
  rename(crs28992res500m = c28992r500, 
         percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)
cbs_500_17 <- st_read(dsn = "C:/Users/kalle/Downloads/cbs_vk500_2017.gpkg") |>
  mutate (year = 2017) |>
  rename(crs28992res500m = c28992r500,
         percentage_geb_nederland_herkomst_nederland = percentage_nederlandse_achtergrond)

cbs_500_16 <- st_read(dsn = "C:/Users/kalle/Downloads/2019-cbs_vk500_2016_v2/CBS_VK500_2016_v2.shp") |>
  mutate (year = 2016) |>
  rename(
    crs28992res500m = C28992R500,
    aantal_inwoners = INWONER,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    percentage_laag_inkomen_huishouden = P_LINK_HH, 
    percentage_hoog_inkomen_huishouden = P_HINK_HH, 
    omgevingsadressendichtheid = OAD, 
    stedelijkheid = STED, 
    geom = geometry
  )

cbs_500_15 <- st_read(dsn = "C:/Users/kalle/Downloads/2019-cbs_vk500_2015_v2/CBS_VK500_2015_v2.shp") |>
  mutate (year = 2015) |>
  rename(
    crs28992res500m = C28992R500,
    aantal_inwoners = INWONER,
    aantal_inwoners_0_tot_15_jaar = INW_014,
    aantal_inwoners_15_tot_25_jaar = INW_1524,
    aantal_inwoners_25_tot_45_jaar = INW_2544,
    aantal_inwoners_45_tot_65_jaar = INW_4564,
    aantal_inwoners_65_jaar_en_ouder = INW_65PL,
    percentage_geb_nederland_herkomst_nederland = P_NL_ACHTG,
    percentage_westerse_migr_achtergr = P_WE_MIG_A,
    percentage_niet_westerse_migr_achtergr = P_NW_MIG_A,
    percentage_laag_inkomen_huishouden = P_LINK_HH, 
    percentage_hoog_inkomen_huishouden = P_HINK_HH, 
    omgevingsadressendichtheid = OAD, 
    stedelijkheid = STED, 
    geom = geometry
  )

cbs_500 <- bind_rows(cbs_500_24, cbs_500_23, cbs_500_22, cbs_500_21, cbs_500_20, cbs_500_19, cbs_500_18, cbs_500_17, cbs_500_16, cbs_500_15) 

rm(cbs_500_24, cbs_500_23, cbs_500_22, cbs_500_21, cbs_500_20, cbs_500_19, cbs_500_18, cbs_500_17, cbs_500_16, cbs_500_15)
gc()

cbs_500 <- cbs_500 |>
  
  select(crs28992res500m, 
          aantal_inwoners,
          aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland,
          percentage_laag_inkomen_huishouden,
          percentage_hoog_inkomen_huishouden, 
          omgevingsadressendichtheid, 
          stedelijkheid, 
          geom, 
          year)|>
  
  relocate(crs28992res500m, geom, year,
          aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland,
          percentage_laag_inkomen_huishouden,
          percentage_hoog_inkomen_huishouden, 
          omgevingsadressendichtheid, 
          stedelijkheid) |>
  
    rename(
          aantal_inwoners_500 = aantal_inwoners,
          aantal_inwoners_0_tot_15_jaar_500 = aantal_inwoners_0_tot_15_jaar,
          aantal_inwoners_15_tot_25_jaar_500 = aantal_inwoners_15_tot_25_jaar,
          aantal_inwoners_25_tot_45_jaar_500 = aantal_inwoners_25_tot_45_jaar,
          aantal_inwoners_45_tot_65_jaar_500 = aantal_inwoners_45_tot_65_jaar,
          aantal_inwoners_65_jaar_en_ouder_500 = aantal_inwoners_65_jaar_en_ouder,
          percentage_geb_nederland_herkomst_nederland_500 = percentage_geb_nederland_herkomst_nederland
  ) |>
  
  mutate(across(where(is.numeric), ~na_if(., -99997))) |>
  mutate(across(where(is.numeric), ~na_if(., -99995)))

```

Add the square data to the pc6 data

```{r}
#match the squares belonging to the postal code area and keep only the square coordinates
df_cbs <- df_pc6 |>
  st_join(cbs_500,join = st_intersects, left = T, largest = T) |>
  select(-c(year.y, aantal_inwoners_500, aantal_inwoners_0_tot_15_jaar_500,
          aantal_inwoners_15_tot_25_jaar_500,
          aantal_inwoners_25_tot_45_jaar_500,
          aantal_inwoners_45_tot_65_jaar_500,
          aantal_inwoners_65_jaar_en_ouder_500,
          percentage_geb_nederland_herkomst_nederland_500, percentage_laag_inkomen_huishouden, percentage_hoog_inkomen_huishouden, omgevingsadressendichtheid, stedelijkheid))

#match the square attributes in the right year
df_cbs <- df_cbs |>
  left_join (st_drop_geometry(cbs_500), by = c("crs28992res500m", "year.x" = "year")) |>
  rename(year = year.x) |>
  mutate(across(where(is.numeric), ~na_if(., NaN)))
```

```{r}
st_write(df_cbs, "./data/raw_df_cbs.gpkg", append = FALSE)
rm(cbs_500, df_pc6)
```

## Step 4: calculate relevant variables and impute missing years

Write function to impute data for missing years

```{r}

df_cbs <- df_cbs |> arrange(postcode_van, year)

impute_ema <- function(df, target_year, history_years, var) {
  imputed <- df |>
    filter(is.na({{ var }}) & year == target_year) |>
    pull(postcode_van) |>
    unique() |>
    sapply(function(pc) {
      hist_values <- df |>
        filter(postcode_van == pc, year %in% history_years) |>
        arrange(year) |>
        pull({{ var }}) |>
        na.omit()
      
      if (length(hist_values) == 0 || all(is.na(hist_values))) {
        NA_real_
      } else if (length(hist_values) == 1) {
        hist_values[1]
      } else {
        tail(EMA(hist_values, 2), 1)
      }
    }) |>
    enframe(name = "postcode_van", value = "imputed_value")
  
  df |>
    left_join(imputed, by = "postcode_van") |>
    mutate({{ var }} := case_when(
      is.na({{ var }}) & year == target_year ~ imputed_value,
      .default = {{ var }}
    )) |>
    select(-imputed_value)
}
```

Impute data:

```{r}
#delete rows with zero residents
df_cbs <- df_cbs |> filter(aantal_inwoners != 0)

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(percentage_hoog_inkomen_huishouden) & is.na(percentage_laag_inkomen_huishouden)))

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddelde_woz_waarde_woning)))

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(gemiddeld_elektriciteitsverbruik_woning)))


df_cbs <- df_cbs |>
  impute_ema(target_year = 2023, history_years = c(2021, 2022), var = percentage_laag_inkomen_huishouden) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = percentage_laag_inkomen_huishouden) |>
  impute_ema(target_year = 2023, history_years = c(2021, 2022), var = percentage_hoog_inkomen_huishouden) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = percentage_hoog_inkomen_huishouden) |>
  impute_ema(target_year = 2015, history_years = c(2017, 2016), var = gemiddelde_woz_waarde_woning) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = gemiddelde_woz_waarde_woning) |>
  impute_ema(target_year = 2024, history_years = c(2022, 2023), var = gemiddeld_elektriciteitsverbruik_woning)

```

Calculate demographic diversity using Herfindahl index

```{r}
##economic diversity
df_cbs <- df_cbs |>
  mutate(
    economic_diversity = 1 - ((percentage_hoog_inkomen_huishouden^2 +
                                 percentage_laag_inkomen_huishouden^2 + 
                                 (100 - percentage_hoog_inkomen_huishouden - percentage_laag_inkomen_huishouden)^2
                                  ) 
         / 10000 
                                 )
         )

summary(df_cbs$economic_diversity)

##ethnic diversity 

df_cbs <- df_cbs |>
  mutate(
    ethnic_diversity = 1 - ((percentage_geb_nederland_herkomst_nederland_500^2 + (100 - percentage_geb_nederland_herkomst_nederland_500)^2) / 10000)
  )

summary(df_cbs$ethnic_diversity)

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(ethnic_diversity)))

##age diversity

#set number in age category to 0 if the total number of people is not missing
df_cbs <- df_cbs |>
  mutate(
    aantal_inwoners_0_tot_15_jaar_500 = if_else(
      is.na(aantal_inwoners_0_tot_15_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_0_tot_15_jaar_500
    ),
    aantal_inwoners_15_tot_25_jaar_500 = if_else(
      is.na(aantal_inwoners_15_tot_25_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_15_tot_25_jaar_500
    ),
    aantal_inwoners_25_tot_45_jaar_500 = if_else(
      is.na(aantal_inwoners_25_tot_45_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_25_tot_45_jaar_500
    ),
    aantal_inwoners_45_tot_65_jaar_500 = if_else(
      is.na(aantal_inwoners_45_tot_65_jaar_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_45_tot_65_jaar_500
    ),
    aantal_inwoners_65_jaar_en_ouder_500 = if_else(
      is.na(aantal_inwoners_65_jaar_en_ouder_500) & !is.na(aantal_inwoners_500),
      0,
      aantal_inwoners_65_jaar_en_ouder_500
    )
  )

#calculate proportions
df_cbs <- df_cbs |>
  mutate(
    percentage_inwoners_0_tot_15_jaar_500 = aantal_inwoners_0_tot_15_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_15_tot_25_jaar_500 = aantal_inwoners_15_tot_25_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_25_tot_45_jaar_500 = aantal_inwoners_25_tot_45_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_45_tot_65_jaar_500 = aantal_inwoners_45_tot_65_jaar_500 / aantal_inwoners_500,
    percentage_inwoners_65_jaar_en_ouder_500 = aantal_inwoners_65_jaar_en_ouder_500 / aantal_inwoners_500,
  )

df_cbs <- df_cbs |>
  mutate(
    age_diversity = 1 - (percentage_inwoners_0_tot_15_jaar_500^2 +
                    percentage_inwoners_15_tot_25_jaar_500^2 +
                    percentage_inwoners_25_tot_45_jaar_500^2 +
                    percentage_inwoners_45_tot_65_jaar_500^2 +
                    percentage_inwoners_65_jaar_en_ouder_500^2)
  )


summary(df_cbs$age_diversity)

df_cbs |> 
  group_by(year) |> 
  summarise(missing_count = sum(is.na(age_diversity)))

```

Add historic solar panel costs from IRENA (<https://www.irena.org/Data/View-data-by-topic/Costs/Solar-costs>)

```{r}
df_cbs <- df_cbs |>
  mutate(solar_cost = case_when(
    year == 2015 ~ NA,
    year == 2016 ~ 0.1291,
    year == 2017 ~ 0.1430,
    year == 2018 ~ 0.1092,
    year == 2019 ~ 0.1040,
    year == 2020 ~ 0.1006,
    year == 2021 ~ 0.0836,
    year == 2022 ~ 0.0912,
    year == 2023 ~ 0.0699,
    year == 2024 ~ NA
  ))

df_cbs <- df_cbs |>
    impute_ema(target_year = 2015, history_years = c(2017, 2016), var = solar_cost) |>
    impute_ema(target_year = 2024, history_years = c(2022, 2023), var = solar_cost)
  
summary(df_cbs$solar_cost)
```

Add electricity price data from CBS (<https://opendata.cbs.nl/#/CBS/nl/dataset/85666NED/table>)

```{r}
## electricity use
summary(df_cbs$gemiddeld_elektriciteitsverbruik_woning)
hist(df_cbs$gemiddeld_elektriciteitsverbruik_woning)

## add electricity price (CBS)
df_cbs <- df_cbs |>
  mutate(electricity_price = case_when(
    #2015
    gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2015 ~ -0.024,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2015 ~ 0.166,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2015 ~ 0.196,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2015 ~ 0.208,
    #2016
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2016 ~ -0.054,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2016 ~ 0.140,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2016 ~ 0.168,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2016 ~ 0.180,
    #2017
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2017 ~ -0.075,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2017 ~ 0.129,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2017 ~ 0.161,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2017 ~ 0.176,
    #2018
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2018 ~ -0.036,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2018 ~ 0.150,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2018 ~ 0.178,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2018 ~ 0.191,
    #2019
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2019 ~ 0.110,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2019 ~ 0.199,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2019 ~ 0.205,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2019 ~ 0.213,
    #2020
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2020 ~ -0.220,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2020 ~ 0.072,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2020 ~ 0.139,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2020 ~ 0.185,
    #2021
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2021 ~ -0.239,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2021 ~ 0.066,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2021 ~ 0.136,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2021 ~ 0.185,
    #2022
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2022 ~ -0.544,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2022 ~ -0.038,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2022 ~ 0.106,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2022 ~ 0.201,
    #2023
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2023 ~ 0.020,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2023 ~ 0.230,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2023 ~ 0.316,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2023 ~ 0.389,
    #2024
        gemiddeld_elektriciteitsverbruik_woning < 1000 & year == 2024 ~ -0.068,
    gemiddeld_elektriciteitsverbruik_woning >= 1000 & gemiddeld_elektriciteitsverbruik_woning < 2500 & year == 2024 ~ 0.174,
    gemiddeld_elektriciteitsverbruik_woning >= 2500 & gemiddeld_elektriciteitsverbruik_woning < 5000 & year == 2024 ~ 0.235,
    gemiddeld_elektriciteitsverbruik_woning >= 5000 & gemiddeld_elektriciteitsverbruik_woning < 15000 & year == 2024 ~ 0.270,
  ))

summary(df_cbs$electricity_price)

## calculate electricity cost
df_cbs <- df_cbs |>
  mutate(electricity_cost = electricity_price * gemiddeld_elektriciteitsverbruik_woning)

summary(df_cbs$electricity_cost)
```

Look at control variables and their missing values

```{r}
## home ownership
summary(df_cbs$percentage_koopwoningen)
summary(df_cbs$percentage_huurwoningen)

df_cbs |> filter(is.na(percentage_koopwoningen) & is.na(percentage_huurwoningen)) |> summarise(count = n())

#unnecessary missings
df_cbs <- df_cbs |>
  mutate(percentage_koopwoningen = if_else (
    is.na(percentage_koopwoningen) & !is.na(percentage_huurwoningen),
    100 - percentage_huurwoningen,
    percentage_koopwoningen
  ))

## ethnicity
summary(df_cbs$percentage_geb_nederland_herkomst_nederland)
hist(df_cbs$percentage_geb_nederland_herkomst_nederland)

## age


#estimating the average age
df_cbs <- df_cbs |>
  mutate(
    age_avg = (
      (7.5 * aantal_inwoners_0_tot_15_jaar) + 
      (20 * aantal_inwoners_15_tot_25_jaar) + 
      (35 * aantal_inwoners_25_tot_45_jaar) +
      (55 * aantal_inwoners_45_tot_65_jaar) +
      (75 * aantal_inwoners_65_jaar_en_ouder)  
              ) 
    / aantal_inwoners
        )

summary(df_cbs$age_avg)
hist(df_cbs$age_avg)
```

### Installed base

Calculating the average energy feed-in in the same 500m x 500m square (weighted by number of households and exluding a streets' own energy feed-in)

```{r}

#calculate the weighted sum of energy feed-in per 500m square per year.
avg_500 <- df_cbs |>
  st_drop_geometry() |>
  group_by(crs28992res500m, year) |>
  summarise(
    w_sum_teruglevering_500 = sum((teruglevering * aantal_part_huishoudens), na.rm = TRUE),
    n_hh_500 = sum(aantal_part_huishoudens),
    .groups = "drop"
  )

#calculate
df_cbs <- df_cbs |> 
  left_join(avg_500, by = c("crs28992res500m", "year")) |>
  mutate(nb_pv = 
           (w_sum_teruglevering_500 - (teruglevering * aantal_part_huishoudens)) / (n_hh_500 - aantal_part_huishoudens)
         )

summary(df_cbs$teruglevering)
summary(df_cbs$nb_pv)

```

## Step 5: Join based on t-1

```{r}
#only select the colums I may use
df_cbs <- df_cbs |>
  select(-c(teruglevering, w_sum_teruglevering_500, n_hh_500, percentage_geb_nederland_herkomst_overig_europa, percentage_geb_nederland_herkomst_buiten_europa, percentage_geb_buiten_nederland_herkomst_europa, percentage_geb_buiten_nederland_herkmst_buiten_europa, percentage_niet_westerse_migr_achtergr, percentage_westerse_migr_achtergr, gemiddelde_huishoudensgrootte, aantal_mannen, aantal_vrouwen, aantal_inwoners_0_tot_15_jaar_500, aantal_inwoners_15_tot_25_jaar_500, aantal_inwoners_25_tot_45_jaar_500, aantal_inwoners_45_tot_65_jaar_500, aantal_inwoners_65_jaar_en_ouder_500, percentage_inwoners_0_tot_15_jaar_500, percentage_inwoners_15_tot_25_jaar_500, percentage_inwoners_25_tot_45_jaar_500, percentage_inwoners_45_tot_65_jaar_500, percentage_inwoners_65_jaar_en_ouder_500))
  
```

Match the predictor data to the solar panel data a year later, given the delay between the decision to adopt solar panels and the installation.

```{r}
#join at t0 if t+1 is not available??

df <- pv |>
  mutate(year_t_minus_one = year - 1) |>
  inner_join(
    df_cbs,
    by = join_by(postcode_van == postcode_van,
                 year_t_minus_one == year)
  ) |>
  rename(id = id.x,
         postcode_tot = postcode_tot.x,
         straatnaam = straatnaam.x) |>
  select( -c(id.y, postcode_tot.y, straatnaam.y, year_t_minus_one)) |>
  relocate(id, postcode_van, postcode_tot, straatnaam, year, teruglevering, nb_pv, percentage_hoog_inkomen_huishouden, percentage_laag_inkomen_huishouden, gemiddelde_woz_waarde_woning, solar_cost, gemiddeld_elektriciteitsverbruik_woning, electricity_price, electricity_cost, economic_diversity, ethnic_diversity, age_diversity, omgevingsadressendichtheid, stedelijkheid, percentage_geb_nederland_herkomst_nederland, age_avg, percentage_koopwoningen, percentage_huurwoningen)
```

Save data!

```{r}

st_write(df, "./data/df_analysis.gpkg", append = F)

```
